---
title: "Recovery Discovery Covariates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{recovery_discovery_covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(MultiStudyNMF)
```

We provide demo data for this example as `data(lymph)`. The R chunk below shows how this data was prepared, using data from PCAWG that can be downloaded [here](https://www.synapse.org/Synapse:syn11726620). As a simple example, we considered two related cancer types (non-Hodgkin's lymphoma and chronic lymphocytic leukemia) from two different cohorts (PCAWG and an additional cohort). For the purposes of demonstrating this approach, we mixed together both cancer types, and separated into two studies by cohort. We chose to use the cancer type as the covariate. We also ordered the mutational categories of this data to match that of the COSMIC reference signatures -- **this is essential for the recovery-discovery sampler.**

```{r,eval=F}
## NOT RUN

# Load in data from two cohorts and re-order to match COSMIC reference
data <- read.csv('../../../Downloads/WGS_PCAWG_2018_02_09/WGS_PCAWG.96.csv')
data(cosmic)
rownames(data) <- paste0(substr(data$Trinucleotide,0,1),'[',
                         data$Mutation.type,']',
                         substr(data$Trinucleotide,3,3))
data <- data[cosmic$Type,]

other <- read.csv('../../../Downloads/WGS_Other_2018_04_13/WGS_Other.96.csv')
rownames(other) <- paste0(substr(other$Trinucleotide,0,1),'[',
                         other$Mutation.type,']',
                         substr(other$Trinucleotide,3,3))
other <- other[cosmic$Type,]

# Subset to the two cancers of interest
cancer_id <- substr(colnames(data),0,10)
names(cancer_id) <- colnames(data)
cancer_id2 <- substr(colnames(other),0,10)
names(cancer_id2) <- colnames(other)

data <- data[,cancer_id%in%c('Lymph.BNHL','Lymph.CLL.')]
cancer_id <- cancer_id[colnames(data)]
other <- other[,cancer_id2%in%c('Lymph.BNHL','Lymph.CLL.')]
cancer_id2 <- cancer_id2[colnames(other)]

# For the purposes of the vignette, randomly subsample both to a smaller size
set.seed(31623)
subsamp1 <- sample(1:ncol(data),40,replace=F)
data <- data[,subsamp1]
cancer_id <- cancer_id[subsamp1]
names(cancer_id) <- NULL

subsamp2 <- sample(which(cancer_id2=='Lymph.CLL.'),30,replace=F)
other <- other[,c(which(cancer_id2=='Lymph.BNHL'),subsamp2)]
cancer_id2 <- cancer_id2[c(which(cancer_id2=='Lymph.BNHL'),subsamp2)]
names(cancer_id2) <- NULL

# Set up final data and covariates input
cancer_id <- ifelse(cancer_id=='Lymph.CLL.',1,0)
cancer_id2 <- ifelse(cancer_id2=='Lymph.CLL.',1,0)
M <- list(as.matrix(data),
          as.matrix(other))
cov <- list(matrix(cancer_id,nrow=1),
            matrix(cancer_id2,nrow=1))
```

Note that the input data is a list of two matrices, each in mutational motifs-by-samples format, and a list of two covariates matrices, each in covariates-by-samples format. Here, both studies have 40 samples.

```{r}
data(lymph)
M <- lymph$M
cov <- lymph$cov
str(M)
str(cov)
```

Now we run the recovery-discovery sampler with covariates.

```{r, results='hide', message=FALSE}
set.seed(62524)
out <- recovery_discovery_cov_fit(M,cov)
```

We can view the signature sharing matrices, which describe which of the estimated signatures are present in each of the two studies, for both the discovery and the recovery components respectively.

```{r}
out$A$discovery
out$A$recovery
```

Here, we see a total of 37 signatures -- 5 in the discovery component (all unique to either one study or the other), and 32 in the recovery component (15 common to both studies, and the rest unique to one or the other). 

We can also view the posterior inclusion probability for each covariate affecting the exposure to each discovery and recovery signature in each study.

```{r}
out$indics$discovery
out$indics$recovery
```

Here, we see that the covariate (cancer type) has a high probability of affecting the exposures to several signatures, including some in the discovery component and some in the recovery component, in each study. 

Similarly, we can view the coefficients for each covariate and discovery or recovery signature in each study. Note that these coefficients represent the effect conditional on inclusion.

```{r}
out$coefs$discovery
out$coefs$recovery
```

For example, in study 2, this output tells us that being a CLL sample results in lower exposures to discovery signature 4 by a factor of `exp(-2.30)` with high certainty (posterior inclusion probability of 0.999 from before), and lower exposures to recovery signature 1 by a factor of `exp(-1.28)` but with lower certainty (posterior inclusion probability of 0.31 from before), among others.

This output also contains the signature estimates for both the discovery and recovery components in `out$P`, as well as the corresponding exposure estimates (`out$E`) and the study-specific normalization factors (`out$W`). More detail about this output can be found in the documentation by viewing `?recovery_discovery_cov_fit`.

Below, we provide two convenience functions to help interpret the output. First, we can annotate each signature by its closest match to the COSMIC v3.2 signatures (further details in `?label_A`) and view the signature sharing matrices as follows for the discovery and recovery components:

```{r}
label_A(out$A$discovery,out$P$discovery)
label_A(out$A$recovery,out$P$recovery)
```

Some signatures may only be present in a small subset of samples. We can view the proportion of samples in each study with relative exposures over a pre-specified threshold as follows (combining both the discovery and the recovery signatures):

```{r}
propA <- proportion_A(out,thresh=0.05,label=T)
propA
```

If we just want to focus on the signature sharing pattern among signatures with high exposures in more than, say, 5\% of samples:

```{r}
propA[,sapply(1:ncol(propA),function(x) max(propA[,x])>0.05)]
```
